---
title: "10. Visualize Phylogenetic Tree"
author:
    - "Will Hannon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
  highlight: tango
  number_sections: no
  theme: default
  toc: yes
  toc_depth: 3
  toc_float:
    collapsed: no
    smooth_scroll: yes
---

The goal of this notebook is to visualize the phylogenetic relationship of the haplotypes identified in the previous notebooks. 

```{r Setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = FALSE)
```

```{r Required Packages, message=FALSE, warning=FALSE, echo=FALSE}

## ==== Install Required Packages ==== ##

## List of all packages needed
packages = c("tidyverse", "RColorBrewer", "ggnetwork", "network", "tidytree", "phytools", "ggtree", "treeio")

## Packages loading
invisible(lapply(c(packages), library, character.only = TRUE))

```

```{r Inputs, echo=T}

## ==== File paths input data ==== ##

if (exists("snakemake")) {

  # Data from lofreq and varscan labeled with subclonal haplotypes
  updated.data = snakemake@input[["1"]]
  
  # Phylogeny 
  phylo.data =  snakemake@input[["2"]]

} else {

  # Data from lofreq and varscan labeled with subclonal haplotypes
  updated.data = "../../results/variants/validated_variants.csv"
  
  # Phylogeny 
  phylo.data = "../../results/phylogeny/haplotype-sequences.fa.treefile"

}

```

## Visualize IQ-TREE Phylogeny

I took the sequences for each of these haplotypes and generated a phylogenetic tree using `iqtree`. The phylogenetic relationship of the clusters was determined by `SPRUCE` with the total set of trees filtered down by bridging reads. The `iqtree` analysis is done in notebook `09-make-phylogenetic-tree.ipynb`.

```{r Import Data, echo=F}

# Assigned SNPs with haplotype labels 
updated.df = read_csv(updated.data, show_col_types = FALSE) %>% 
  # For consistency, change the names of G1, G2, and G-1-1
  mutate(Haplotype = case_when(
    Haplotype == "genome-1" ~ "genome 01",
    Haplotype == "genome-1-1" ~ "genome 1",
    Haplotype == "genome-2" ~ "genome 2",
    T ~ Haplotype
  )) 

# Get the background of the haplotypes
background.label = updated.df %>% 
  select(Haplotype, Background) %>% 
  filter(!Haplotype %in% c("subclonal", "genome 2", "genome 01", "genome 1", "both")) %>% 
  mutate(Haplotype = str_replace(Haplotype, " ", "_")) %>% 
  distinct()

# Mutation counts 
mutation.counts = updated.df %>% 
  filter(!Haplotype %in% c('subclonal', 'fixed', 'both')) %>% 
  select(POS, REF, ALT, Haplotype) %>% 
  distinct() %>% 
  group_by(Haplotype) %>% 
  count() %>%
  ungroup() %>% 
  rename(label = Haplotype, mutation_count = n)

# Phylogenetic tree 
tree = read.iqtree(phylo.data)

```

```{r Root the Tree, echo=T}

# Name of the out group for rooting the tree
root = "SSPE_ancestor"

# Root the tree
tree@phylo = root(tree@phylo, outgroup = root)

# Add the background labels
tree.df = left_join(as_tibble(tree), background.label, by=c("label" = "Haplotype")) 

# Get the nodes in the tree
genome.1.node = MRCA(tree, filter(background.label, Background == "genome-1")$Haplotype)
genome.2.node = MRCA(tree, filter(background.label, Background == "genome-2")$Haplotype)
genome.1.offspring = offspring(tree, genome.1.node)
genome.2.offspring = offspring(tree, genome.2.node)

# Add the background labels
tree.df = as_tibble(tree) %>% 
  mutate(branch.length = if_else(UFboot < 70 & !is.na(UFboot), 0.0, branch.length)) %>% 
  mutate(Background = case_when(node %in% genome.1.offspring ~ "genome-1",
                                node %in% genome.2.offspring ~ "genome-2",
                                TRUE ~ "ancestral")) %>% 
  mutate(label = str_replace(label, "_", " "))  %>% 
  left_join(., mutation.counts, by = "label")

```

```{r Plot the Tree, fig.align="center", fig.width=32, fig.height=15}

tree.df %>% 
  as.treedata() %>% 
  ggtree(size = 3, aes(color = Background)) +
    geom_point2(aes(subset=(node==1)), size=12, color="#4d4d4d") + 
    geom_point2(aes(subset=(node==genome.1.node)), size=12, color="#307EC2") + 
    geom_point2(aes(subset=(node==genome.2.node)), size=12, color="#B63F3E") + 
    geom_tiplab(hjust = -.05, size = 8) + 
    scale_color_manual(values = c("#4d4d4d", "#307EC2", "#B63F3E")) + 
    theme(legend.position = "none") 

# ggsave("../../results/phylogeny/sspe-phylogeny.png", width = 32, height = 15, dpi = 300)

```


