---
title: "2. Cluster Major Genotypes"
author:
    - "Will Hannon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
  highlight: tango
  number_sections: no
  theme: default
  toc: yes
  toc_depth: 3
  toc_float:
    collapsed: no
    smooth_scroll: yes
---

As observed by the Cattaneo lab, there are clearly sets of mutations that move together in frequency across every spatial region of the brain. In particular, there are two sets of mutations who's average frequency appears to add up to 100% in every tissue. These two 'genotypes' appear to be the background upon which all other mutations arose. 

The goal of this notebook is to formalize an approach to phase these mutations by clustering variants that have strongly correlated frequencies across tissue specimens using  k-medoids clustering applied to a correlation matrix. 

In this notebook, I identify the 'genotypes' `Genome 1` and `Genome 2`. In addition to these two 'genotypes', I identify a third set of mutations that appears to arise on `Genome 1`, but is at very low frequency in the Frontal Cortex 2 specimen.

```{r Setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = FALSE)
```

```{r Required Packages, message=FALSE, warning=FALSE}

## ==== Install Required Packages ==== ##

## List of all packages needed
packages = c("tidyverse", "ComplexHeatmap", "cluster", "knitr", "ggrepel")

## Packages loading
invisible(lapply(c(packages), library, character.only = TRUE))

```

The inputs for this analysis are (1) the filtered and combined variants from `quality-control.Rmd` and (2) annotations of the positions of the main Measles open reading frames, and (3) the mutations added to the patient-specific reference sequencing. 

```{r Input, echo=T}

## ==== File paths input data ==== ##

if (exists("snakemake")) {
  # Variants that were filtered and corrected in `quality-control.Rmd`
  variant.data = snakemake@params[['incsv']]
  # Annotations for the standard Measles ORFs ~ for making pretty figures.
  annotations.filepath = snakemake@params[['annotations']]
  # Path to the ancestral mutations
  ancestral.data = snakemake@params[['ancestral']]
} else {
  # Variants that were filtered and corrected in `quality-control.Rmd`
  variant.data = "../../results/variants/filtered_variants.csv"
  # Annotations for the standard Measles ORFs ~ for making pretty figures.
  annotations.filepath = "../../config/annotations.csv"
  # Path to the ancestral mutations
  ancestral.data = "../../config/ref/annotated_SSPE_consensus_snps.csv"
}

```

The output of this analysis are the filtered variants annotated by whether they are `Genome-1`, `Genome-2`, `Genome-1-1`, or `subclonal` to these three 'genotypes'.

```{r Outputs, echo=T}

## ==== File paths output data ==== ##

# Path to save output
if (exists("snakemake")) {
  output.path = snakemake@params[['outcsv']]
} else {
  output.path = "../../results/variants/genotyped_variants.csv"
}

```

## Clusters of Variants

This is what the data we have looks like. 

```{r Read in Data, echo=F, warning=F, message=F}

selected.variants.df = read_csv(variant.data, show_col_types = FALSE)

head(selected.variants.df) %>% 
  kable(., caption = "Filtered Variants")

```
It's clear that there groups of mutations that are found at the same frequency across tissues. This was originally observed in a pilot tissue specimen that was chipped from the frozen brain. This sample is called `SSPE 1`. 

```{r Annotations, message=FALSE, warning=FALSE, echo=FALSE}

# Import genome annotations
annotations.df = read_csv(annotations.filepath, show_col_types = FALSE)

# Get the interval positions for these genes 
N = annotations.df %>% dplyr::filter(Locus == "N")
P.V.C = annotations.df %>% dplyr::filter(`Protein Name` == "phosphoprotein")
M = annotations.df %>% dplyr::filter(Locus == "M")
F. = annotations.df %>% dplyr::filter(Locus == "F")
H = annotations.df %>% dplyr::filter(Locus == "H")
L = annotations.df %>% dplyr::filter(Locus == "L")

# Add gene names to the variants
selected.variants.df = selected.variants.df %>% 
  mutate(Gene = case_when(POS < N$Start ~ "3'UTR",
                          POS > L$Stop ~ "5'UTR", 
                          TRUE ~ Gene_Name)) 

tissue_colors = c("#ef476f","#ffd166","#06d6a0","#118ab2","#073b4c","#DDA0DD")

# Tissue order - roughly by the location in the brain 
tissue_order = c(
  "SSPE 1", 
  "SSPE 2",
  "Frontal Cortex 2", 
  "Frontal Cortex 1", 
  "Frontal Cortex 3", 
  "Parietal Lobe",
  "Temporal Lobe", 
  "Occipital Lobe",
  "Hippocampus",
  "Internal Capsule", 
  "Midbrain", 
  "UBS",
  "Brain Stem",
  "Cerebellum",
  "Cerebellum Nucleus"
)

```

We can see these mutations clearly if we draw 'gates' of frequency around mutations on the plot. Below, I'm drawing 'gates' around 60-75% and 30-45% of the genome respectively.

```{r Figure 2 Data, echo=T, warning=FALSE}

# Read in the consensus SNPs
ancestral.df = read_csv(ancestral.data, show_col_types = FALSE) %>% 
  mutate(SNP = paste0(REF, POS, ALT)) %>% 
  select(POS, REF, ALT, SNP) %>% 
  mutate(AF = 1.0) 

# Format the data for plotting
sspe.1.df = selected.variants.df %>% 
  # Select the pilot specimen
  filter(Tissue == "SSPE 1") %>% 
  # Select the important columns 
  select(colnames(ancestral.df)) %>% 
  # Add in the ancestral SNPs 
  bind_rows(., ancestral.df) %>% 
  # Bind variants by frequency
  mutate(Genotype = case_when(
    AF >= 0.9 ~ "CBA",
    AF >= 0.6 & AF <= 0.75 ~ "CG1", 
    AF >= 0.3 & AF <= 0.45 ~ "CG2",
    TRUE ~ "subclonal"
  )) %>% 
  # Point size should be small for mutation we don't know about
  mutate(Size = if_else(Genotype == "subclonal", "small", "normal")) %>% 
  # Point shape depends on whether the mutation is a transition or transversion
  mutate(Shape = case_when(
    REF == "A" & ALT == "G" ~ "A to G / U to C",
    REF == "T" & ALT == "C" ~ "A to G / U to C",
    REF == "G" & ALT == "A" ~ "Other Transitions",
    REF == "C" & ALT == "T" ~ "Other Transitions",
    TRUE ~ "Transversions"
  )) %>% 
  # We want to highlight mutations that are candidate 'Drivers' of spread
  mutate(Driver = case_when(
    SNP == "G3812A" ~ "M-W125*",
    SNP == "T6817A" ~ "F-L454M",
    SNP == "T3586C" ~ "M-F50S",
    SNP == "T7293C" ~ "H-I8T",
    SNP == "C7036T" ~ "F-Q527*",
    TRUE ~ ""
  )) 
  
# Binned SNPS (i.e. CBA, CG1, and CG2)
binnned.snps = sspe.1.df %>% 
  filter(Genotype %in% c("CBA", "CG1", "CG2") & Driver == "")
# Un-binned SNPS (i.e. subclonal)
subclonal.snps = sspe.1.df %>% 
  filter(Genotype == 'subclonal' & Driver == "")
# Putative driver SNPs
driver.snps = sspe.1.df %>% 
  filter(Driver != "")

sspe.1.df %>% 
  filter(Genotype == "CG2") %>% 
  pull(SNP) %>% 
  unique() %>% 
  length()

```

```{r Figure 2 plot, fig.align='center', fig.width=10, fig.height=5}

# Make Figure 2 of the paper
ggplot(data = binnned.snps, aes(x = POS, y = AF, color = Genotype, shape = Shape)) +
  geom_point(size = 2) + 
  geom_point(data = subclonal.snps, color = "#878484", size = 1) + 
  geom_point(data = driver.snps, color = "black", size = 3) + 
  geom_text_repel(data = driver.snps, aes(label = Driver), color = "black", point.padding = 1) + 
  xlab("Position") + 
  ylab("Allele Frequency") +
  scale_color_manual(values = c("#d9cb34", "#307EC2", "#B63F3E")) +
  scale_shape_discrete(name="Mutation") +
  
  # == Annotations of genes == #
  
  # 3' Start
  annotate(geom = "text", x = (N$Start - 150), y =  -.05, label = "3'", size = 5) +
   
  # Nucleoprotein
  annotate("rect", xmin = N$Start-1, xmax = N$Stop, ymin = -.1, ymax = 0 ,
             alpha = .5, fill = "#d97634", col = "black", size = .1) +
  annotate(geom = "text", x = (N$Stop + N$Start)/2, y =  -.05, label = "N", size = 5) +
  annotate("rect", xmin = N$Start-1, xmax = N$Stop, ymin = 0, ymax = 1.1,
                 alpha = .05, fill = "#d97634", size = .1) +  
 
  # P/V/C
  annotate("rect", xmin = P.V.C$Start, xmax = P.V.C$Stop, ymin = -.1, ymax = 0 ,
             alpha = .5, fill = "#fcff3d", col = "black", size = .1) +
  annotate(geom = "text", x = (P.V.C$Stop + P.V.C$Start)/2, y =  -.05, label = "P/V/C", size = 5) +
  annotate("rect", xmin = P.V.C$Start, xmax = P.V.C$Stop, ymin = 0, ymax = 1.1,
             alpha = .05, fill = "#fcff3d", size = .1) +
 
  # Matrix protein
  annotate("rect", xmin = M$Start, xmax = M$Stop, ymin = -.1, ymax = 0 ,
             alpha = .5, fill = "#0db51b", col = "black", size = .1) +
  annotate(geom = "text", x = (M$Stop + M$Start)/2, y =  -.05, label = "M", size = 5)  +
  annotate("rect", xmin = M$Start, xmax = M$Stop, ymin = 0, ymax = 1.1,
             alpha = .05, fill = "#0db51b", size = .1) +
 
  # Fusion protein
  annotate("rect", xmin = F.$Start, xmax = F.$Stop, ymin = -.1, ymax = 0 ,
             alpha = .5, fill = "#6A3D9A", col = "black", size = .1) +
  annotate(geom = "text", x = (F.$Stop + F.$Start)/2, y =  -.05, label = "F", size = 5) +
  annotate("rect", xmin = F.$Start, xmax = F.$Stop, ymin = 0, ymax = 1.1,
             alpha = .05, fill = "#6A3D9A", size = .1) +

  # Hemagluttinin
  annotate("rect", xmin = H$Start, xmax = H$Stop, ymin = -.1, ymax = 0 ,
             alpha = .5, fill = "#0251d9", col = "black", size = .1) +
  annotate(geom = "text", x = (H$Stop + H$Start)/2, y =  -.05, label = "H", size = 5) +
  annotate("rect", xmin = H$Start, xmax = H$Stop, ymin = 0, ymax = 1.1,
             alpha = .05, fill = "#0251d9", size = .1) +

  # Large protein
  annotate("rect", xmin = L$Start, xmax = L$Stop, ymin = -.1, ymax = 0 ,
             alpha = .5, fill = "#bf1515", col = "black", size = .1) +
  annotate(geom = "text", x = (L$Stop + L$Start)/2, y = -.05, label = "L", size = 5) +
  annotate("rect", xmin = L$Start, xmax = L$Stop, ymin = 0, ymax = 1.1,
             alpha = .05, fill = "#bf1515", size = .1) +
  
  # 5' Start
  annotate(geom = "text", x = (L$Stop + 150), y =  -.05, label = "5'", size = 5) +
 
  
  # Themes
  theme_classic() +
  theme(panel.grid.major.y = element_line(colour="grey", linetype="dashed")) +
  theme(legend.position="bottom", legend.box = "horizontal") +
  theme(legend.box.background =element_blank()) +
  theme(text=element_text(size=16)) + 
  theme(legend.text = element_text(size=12)) +
  theme(axis.text.x = element_text(size=16)) + 
  theme(axis.title.x = element_text(size=16)) + 
  scale_x_continuous(limits=c(-150, 16150), breaks=c(0, 5000, 10000, 15000)) +
  scale_y_continuous(breaks=c(0, 0.25, 0.5, 0.75, 1.0)) +
  theme(legend.position = "bottom") 

ggsave("./figure-2.png", width = 11, height = 5, dpi = 300)

```

What are the counts of mutations in each of these three bins (`CBA`: "Candidate Brain Ancestor", `CG1`: "Candidate Genome 1", and `CG2`: "Candidate Genome 2")?

```{r Mutations in Each Genotype}
# What's the count of SNPs in each category?
sspe.1.df %>% 
  group_by(Genotype) %>% 
  count() %>% 
  kable(., caption = "Candidate Genotype Variant Counts")


sspe.1.df %>% 
  filter(Genotype == "CG2") %>% 
  pull(SNP) %>% 
  unique() %>% 
  paste(., collapse = ", ")

```

These observations are what spurred the following approach to formalize a way to phase the mutations in these genotypes using all 15 tissues. 

## Assign Mutations to Major Backgrounds

We know from experiments and preliminary observations that there are **at least two sets of mutations present, but not fixed in every tissue sample**. These two sets of mutations are on the background of all viral sequences from this patient. Since these two sets of mutations are in every sample and their frequencies are anti correlated, we should be able to identify these mutations by clustering them based on their co/anti-correlation. 

```{r Shared Mutations, warning = F, message = F, echo=T}

# All thirteen of the tissue samples 
tissues = selected.variants.df %>% 
  pull(Tissue) %>% 
  unique()

# Get the mutations in every single tissue
all.mutations = selected.variants.df %>% 
  count(SNP) %>%
  filter(n == length(tissues)) %>%
  pull(SNP)

print(paste("There are", length(all.mutations), "mutations in every tissue sample"))

```

Although the reference seems to capture nearly every fixed mutation, there are four that aren't included in the reference. These are all in one continuous four nucleotide stretch in Nucleoprotein (N). I don't know why these don't get captured in the reference assembly, but it likely has to do with issues with the alignment of these reads. **I'll exclude these from the downstream analysis and add these back into the SSPE reference.** 

```{r Remaining Fixed Mutations, warning=F, message=F, echo=T}

# Remove the fixed mutations and get the long form of mutations in all samples. 
fixed.in.all = selected.variants.df %>% 
  filter(SNP %in% all.mutations) %>%
  filter(AF >= 0.95) %>% 
  group_by(SNP) %>% 
  count() %>% 
  filter(n == length(tissues)) %>%
  pull(SNP)

print(paste("There are", length(fixed.in.all), "fixed mutations in every tissue sample that aren't in the reference."))
print(paste("Fixed but not in reference:", fixed.in.all))
```

First, I'll make a correlation matrix between all of the SNP frequencies for SNPs that are found in every tissue and aren't fixed in every tissue (i.e. in the reference). 

```{r SNP Correlation, warning=F, message=F, echo=T, fig.align='center', fig.width=11, fig.height=9}

# Convert these mutations to long form where each column is a SNP and the row is it's frequency.
frequency.by.column = selected.variants.df %>% 
  filter(SNP %in% all.mutations) %>%
  filter(!SNP %in% fixed.in.all) %>% 
  select(AF, SNP, Tissue) %>% 
  pivot_wider(names_from = SNP, values_from = AF, values_fill = 0) %>% 
  select(!Tissue) 

# Calculate R between every pair of columns handling NAs as a correlation matrix
SNP.correlation = cor(frequency.by.column, use="pairwise.complete.obs")

Heatmap(SNP.correlation , 
        rect_gp = gpar(col = "white", lwd = 1),
        column_names_gp = grid::gpar(fontsize = 10),
        row_names_gp = grid::gpar(fontsize = 10),
        show_column_dend = FALSE, 
        show_row_dend = FALSE)
```

Most are strongly correlated (or anti correlated), but some aren't so correlated with one another. Below is a histogram showing this more clearly. I'll filter out SNPs that aren't so well correlated with one another to make it easier to cluster the mutations on `Genome-1` and `Genome-2`. 

```{r Correlation Histogram, echo=T, fig.align='center', fig.width=6, fig.height=4}

# Get the R-squared from the correlation matrix
SNP.correlation.squared = SNP.correlation^2

# Plot the histogram
strong.corr.df = data.frame(SNP = names(rowMeans(SNP.correlation.squared)), Corr_Mean = rowMeans(SNP.correlation.squared))

strong.corr.df %>% 
  ggplot(aes(x = Corr_Mean)) +
    geom_histogram(bins = 40) +
    geom_vline(xintercept = 0.75, color = "red", linetype = 2) +
    xlab("R-squared") + 
    ylab("Count") +
    theme_bw(15)

```

```{r Filter Correlation, echo=T, warning=F, message=F}

# Take only SNPs with greater than .75 based on the above histogram
filtered.SNPs = rowMeans(SNP.correlation.squared)[rowMeans(SNP.correlation.squared) >= 0.75]

# Remove the fixed mutations and get the long form
filtered.SNPs.df = selected.variants.df %>% 
  filter(!SNP %in% fixed.in.all) %>% 
  filter(SNP %in% names(filtered.SNPs)) %>% 
  select(AF, SNP, Tissue) %>% 
  pivot_wider(names_from = SNP, values_from = AF, values_fill = 0) %>% 
  select(!Tissue) 
  
# Calculate R between every pair of columns handling NAs 
filtered.SNP.correlation = cor(filtered.SNPs.df, use="pairwise.complete.obs")

# Convert to distrance (postive corr is close to 0)
filtered.SNP.dist = as.dist(1 - filtered.SNP.correlation)

# Create k-medoids clustering with 3 clusters
filtered.SNP.kmedoids = pam(filtered.SNP.dist, 3) 
kclusters = filtered.SNP.kmedoids$cluster

kclusters.df = data.frame(SNP = names(kclusters), cluster = kclusters)

# Assign the clusters to the original data frame
kmedoids.SNPs = selected.variants.df %>% 
  filter(!SNP %in% fixed.in.all) %>% 
  filter(SNP %in% all.mutations) %>% 
  left_join(., kclusters.df, by = "SNP") %>% 
  mutate(cluster = if_else(is.na(cluster), "no cluster", as.character(cluster)))

# Mean of each tissue
tissue.mean = kmedoids.SNPs %>% 
  group_by(Tissue, cluster) %>% 
  summarize(AF = mean(AF))

```
What's interesting, is that these mutations aren't best explained with 2 clusters (i.e. `Genome-1` and `Genome-2`) but rather with *3 clusters*! 

```{r Plot the clusters, fig.align='center', fig.width=14, fig.height=10}

kmedoids.SNPs %>% 
  ggplot(aes(x = reorder(Tissue, -AF), y = AF)) +
    geom_line(aes(group = SNP)) +
    geom_line(data = tissue.mean, aes(x = Tissue, y = AF, group = 1, col = (cluster)), size = 1) +
    facet_wrap(~cluster) + 
    xlab("Tissue") + 
    ylab("Allele Frequency") +
    theme_bw(20) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    theme(legend.position = "none") 

# ggsave("../../results/figures/each-mutation-cluster.png", width = 14, height = 10, dpi = 300)

```

It's clear from the above plot that mutations grouped in clusters have very similar frequencies across tissues as expected. What's interesting is that clusters 2 and 3 are found at very similar frequencies in every tissue but the `Frontal Cortex 2` sample. It is likely that these are an early divergence of one of the major genotypes.

We'll call cluster 3 `Genome-1`, cluster 2 `Genome-1-1`, and cluster 1 `Genome-2`. The 'no cluster' mutations are simply mutations found in every single tissue that don't neatly fit on `Genome-1`,`Genome-2`, or `Genome-1-1`. These will take some more work to resolve.

```{r Extract G1/G2, warning=F, message=F, echo=T}

g1.SNPs = kmedoids.SNPs %>% 
  filter(cluster == 3) %>% 
  pull(SNP)

sub.g1.SNPs = kmedoids.SNPs %>% 
  filter(cluster == 2) %>% 
  pull(SNP)

g2.SNPs = kmedoids.SNPs %>% 
  filter(cluster == 1) %>% 
  pull(SNP)

labeled.df = selected.variants.df %>% 
  filter(!SNP %in% fixed.in.all) %>% # excluding these four odd 'fixed mutations'. These are better explored in IGV
  mutate(Background = case_when(SNP %in% g1.SNPs ~ "genome-1", 
                                SNP %in% g2.SNPs ~ "genome-2",
                                SNP %in% sub.g1.SNPs ~ "genome-1",
                                TRUE ~ "subclonal")) %>% 
  mutate(Haplotype = case_when(SNP %in% g1.SNPs ~ "genome-1", 
                               SNP %in% g2.SNPs ~ "genome-2",
                               SNP %in% sub.g1.SNPs ~ "genome-1-1",
                               TRUE ~ "subclonal"))
```

I'll write out the newly annotated variant dataframe for further analyses. 

```{r Write Out, warning=F, message=F, echo=T}

# Write these out to a file 
print(paste("Writing the results to", output.path))
write_csv(labeled.df, output.path)

```

## Exploring Genome-1 and Genome-2 

These two clusters of mutations (`Genome-1` and `Genome-2`) are ancestral to all other sequences in the brain. We can see that by summing their frequency in every compartment and seeing that they add up to 1. 

```{r G1 vs. G2 combined, fig.align='center', fig.width=10, fig.height=8, warning=F, message=F}

genotype_order = c("genome-2", "genome-1")
genome.colors = c("#307EC2", "#B63F3E")

genotype.plot.df = labeled.df %>% 
  filter(Haplotype %in% c("genome-1", "genome-2"))  %>% 
  select(Haplotype, Tissue, AF) %>% 
  group_by(Haplotype, Tissue) %>% 
  summarize(AF = mean(AF))

padding = genotype.plot.df %>% 
  group_by(Tissue) %>% 
  summarize(Padding = (1 - sum(AF))/2)

genotype.plot.df = genotype.plot.df %>% 
  left_join(., padding) %>% 
  mutate(AF = AF + Padding) %>% 
  select(!Padding)

genotype.plot.df$Haplotype = factor(genotype.plot.df$Haplotype, level = genotype_order)
  
genotype.plot.df %>% 
  ggplot(aes(x = factor(Tissue, levels = tissue_order), y = AF, fill = Haplotype, group = Haplotype)) + 
  geom_area() +
    xlab("Tissue") +
    ylab("Allele Frequency") +
    scale_fill_manual(values = genome.colors) +
    theme_bw(20) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    theme(plot.title = element_text(hjust = 0.5))

```

Now, let's revisit the distribution of mutations in the genome colored by the mutations that have been identified as `Genome-1` or `Genome-2`. We can see how these mutations are distributed in the brain. 

```{r G1/G2 Distribution, warning=F, message=F, fig.align='center', fig.width=18, fig.height=12}

labeled.df %>% 
  filter(Background %in% c('genome-1', 'genome-2')) %>% 
  ggplot(aes(x = POS, y = factor(Tissue, levels = rev(tissue_order)))) +
    geom_point(aes(col = Background), shape = 108, size = 6) +
    xlab("Position") +
    # Annotations of genes #
    # Nucleoprotein
    annotate("rect", xmin = N$Start-1, xmax = N$Stop, ymin = 0, ymax = .5 ,
               alpha = .2, fill = "#d97634", col = "black", size = .1) +
    annotate(geom = "text", x = (N$Stop + N$Start)/2, y = .25, label = "N", family = "helvetica", size = 10) +
    annotate("rect", xmin = N$Start-1, xmax = N$Stop, ymin = .5, ymax = 16 ,
                   alpha = .1, fill = "#d97634", size = .1) +  
    # P/V/C
    annotate("rect", xmin = P.V.C$Start, xmax = P.V.C$Stop, ymin = 0, ymax = .5 ,
               alpha = .2, fill = "#fcff3d", col = "black", size = .1) +
    annotate(geom = "text", x = (P.V.C$Stop + P.V.C$Start)/2, y = .25, label = "P/V/C", family = "helvetica", size = 10) +
    annotate("rect", xmin = P.V.C$Start, xmax = P.V.C$Stop, ymin = 0, ymax = 16 ,
               alpha = .1, fill = "#fcff3d", size = .1) +
    # Matrix protein
    annotate("rect", xmin = M$Start, xmax = M$Stop, ymin = 0, ymax = .5 ,
               alpha = .2, fill = "#0db51b", col = "black", size = .1) +
    annotate(geom = "text", x = (M$Stop + M$Start)/2, y = .25, label = "M", family = "helvetica", size = 10)  +
    annotate("rect", xmin = M$Start, xmax = M$Stop, ymin = 0, ymax = 16 ,
               alpha = .1, fill = "#0db51b", size = .1) +
    # Fusion protein
    annotate("rect", xmin = F.$Start, xmax = F.$Stop, ymin = 0, ymax = .5 ,
               alpha = .2, fill = "#6A3D9A", col = "black", size = .1) +
    annotate(geom = "text", x = (F.$Stop + F.$Start)/2, y = .25, label = "F", family = "helvetica", size = 10) +
    annotate("rect", xmin = F.$Start, xmax = F.$Stop, ymin = 0, ymax = 16 ,
               alpha = .1, fill = "#6A3D9A", size = .1) +
    # Hemagluttinin
    annotate("rect", xmin = H$Start, xmax = H$Stop, ymin = 0, ymax = .5 ,
               alpha = .2, fill = "#0251d9", col = "black", size = .1) +
    annotate(geom = "text", x = (H$Stop + H$Start)/2, y = .25, label = "H", family = "helvetica", size = 10) +
    annotate("rect", xmin = H$Start, xmax = H$Stop, ymin = 0, ymax = 16,
               alpha = .1, fill = "#0251d9", size = .1) +
    # Large protein
    annotate("rect", xmin = L$Start, xmax = L$Stop, ymin = 0, ymax = .5 ,
               alpha = .2, fill = "#bf1515", col = "black", size = .1) +
    annotate(geom = "text", x = (L$Stop + L$Start)/2, y = .25, label = "L", family = "helvetica", size = 10) +
    annotate("rect", xmin = L$Start, xmax = L$Stop, ymin = 0, ymax = 16 ,
               alpha = .1, fill = "#bf1515", size = .1) +
    scale_color_manual(values = c("#307EC2", "#B63F3E", "grey")) +
    theme_classic() +
    theme(panel.grid.major.y = element_line(colour="grey", linetype="dashed")) +
    theme(legend.position="bottom", legend.box = "horizontal") +
    theme(legend.box.background = element_rect(colour = "black")) +
    theme(text=element_text(size=24)) + 
    theme(axis.text.x = element_text(size=28)) + 
    theme(axis.title.x = element_text(size=30)) + 
    theme(plot.title = element_text(hjust = 0.5))  +
    theme(axis.title.y = element_blank()) +
    scale_x_continuous(limits=c(0,16000)) +
    theme(legend.position = "bottom") 

# ggsave("../../results/figures/snps-in-g1-and-g2.png", width = 18, height = 12, dpi = 300)

```
Interestingly, only `Genome-1` has mutation in the Matrix protein. This is kind of surprising given how hypermutated the Matrix protein is. 

Overall, there are more mutation in `Genome-1` than there are in `Genome-2`, but this is mostly driven by this large block of mutations in the Matrix. 

```{r Mutations per genotype}

# All mutations 
all.mutations.included = labeled.df %>%
  filter(Background %in% c('genome-1', 'genome-2')) %>% 
  select(Background, SNP) %>% 
  distinct() %>% 
  group_by(Background) %>% 
  count()

# Excluding the mutations in Matrix
excluding.only.M = labeled.df %>%
  filter(Background %in% c('genome-1', 'genome-2')) %>% 
  filter(Gene_Name != "M") %>%
  select(Background, SNP) %>% 
  distinct() %>% 
  group_by(Background) %>% 
  count()

print(paste("There are",
            filter(all.mutations.included, Background == "genome-1")$n,
            "mutations in genome-1 and",
            filter(all.mutations.included, Background == "genome-2")$n,
            "mutations in genome-2"))

print(paste("There are",
            filter(excluding.only.M, Background == "genome-1")$n,
            "mutations in genome-1 and",
            filter(excluding.only.M, Background == "genome-2")$n,
            "mutations in genome-2 when you exclude Matrix"))

```

This is very important observation about the haplotypes since Matrix likely accumulates mutations at a different rate than the rest of the genome since there is no selective pressure to maintain the function of Matrix. 

## Which mutations don't cluster? 

Some mutations that are present in every tissue aren't correlating well with either the `Genome-1`, `Genome-2`, or `Genome-1-1` mutations. Let's take a look at these 'no cluster` mutations. 

```{r No Cluster Mutations, warning=F, message=F, fig.align='center', fig.width=12, fig.height=6}

tissue.mean = labeled.df %>% 
  filter(Haplotype %in% c("genome-1", "genome-2")) %>% 
  group_by(Tissue, Haplotype) %>% 
  summarize(AF.mean = mean(AF, na.rm = TRUE), 
            SD = sd(AF, na.rm = TRUE),
            N = n()) %>% 
  mutate(SE = SD / sqrt(N),
         Lower.CI = qt(1 - (0.05 / 2), N - 1) * SE,
         Upper.CI = qt(1 - (0.05 / 2), N - 1) * SE) %>% 
  rename("AF" = AF.mean)


kmedoids.SNPs %>% 
  filter(cluster == 'no cluster') %>% 
  mutate(title = paste(Gene_Name, AA_Change, sep = "-")) %>% 
  ggplot(aes(x = factor(Tissue, levels = tissue_order), y = AF, group = SNP )) + 
    geom_ribbon(data = tissue.mean, aes(x=factor(Tissue, levels = tissue_order),
                                        ymin=AF-SD,
                                        ymax=AF+SD,
                                        group = Haplotype,
                                        fill = Haplotype), alpha=0.2, colour = NA) +
    geom_line() + 
    scale_fill_manual(values=c("#424ef5", "#cf1919")) + 
    xlab("Tissue") + 
    ylab("Allele Frequency") +
    facet_wrap(~title) + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

```

Three of these are particularly interesting, a stop codon towards the terminus of F, a mutation in the C-terminal of H, and a missense mutation in the N protein. 

```{r Interesting Mutations, warning=F, message=F, fig.align='center', fig.width=12, fig.height=4}

tissue.mean = labeled.df %>% 
  filter(Haplotype %in% c("genome-1", "genome-2")) %>% 
  group_by(Tissue, Haplotype) %>% 
  summarize(AF.mean = mean(AF, na.rm = TRUE), 
            SD = sd(AF, na.rm = TRUE),
            N = n()) %>% 
  mutate(SE = SD / sqrt(N),
         Lower.CI = qt(1 - (0.05 / 2), N - 1) * SE,
         Upper.CI = qt(1 - (0.05 / 2), N - 1) * SE) %>% 
  rename("AF" = AF.mean)

kmedoids.SNPs %>% 
  filter(AA_Change %in% c("Gln527*", "Ile8Thr", "Ala235Thr")) %>% 
  mutate(title = paste(Gene_Name, AA_Change, sep = "-")) %>% 
  ggplot(aes(x = factor(Tissue, levels = tissue_order), y = AF, group = SNP )) + 
    geom_ribbon(data = tissue.mean, aes(x=factor(Tissue, levels = tissue_order),
                                        ymin=AF-SD,
                                        ymax=AF+SD,
                                        group = Haplotype,
                                        fill = Haplotype), alpha=0.2, colour = NA) +
    geom_line() + 
    scale_fill_manual(values=c("#424ef5", "#cf1919")) + 
    xlab("Tissue") + 
    ylab("Allele Frequency") +
    facet_wrap(~title) + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

# ggsave("../../results/figures/snps-in-both.png", width = 8, height = 4, dpi = 300)

```

