---
title:"6. Prepare Spruce Input"
author:
    - "Will Hannon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
  highlight: tango
  number_sections: no
  theme: default
  toc: yes
  toc_depth: 3
  toc_float:
    collapsed: no
    smooth_scroll: yes
---

The goal of this notebook is look at how the haplotype frequencies add up in each compartment. In addition to this, we want to prepare the data for SPRUCE to see if we can use this frequency information to fit plausible phylogenetic trees. 

```{r Setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = FALSE)
```

```{r Required Packages, message=FALSE, warning=FALSE, echo=FALSE}

## ==== Install Required Packages ==== ##

## List of all packages needed
packages = c("tidyverse", "foreach", "emdbook", "RColorBrewer", "data.table", "ggnetwork", "network")

## Packages loading
invisible(lapply(c(packages), library, character.only = TRUE))

```

```{r Inputs, echo=T}

## ==== File paths input data ==== ##

if (exists("snakemake")) {

  # Data from lofreq and varscan labeled with subclonal haplotypes
  updated.data = snakemake@params[["incsv"]]

  # Depth per position
  merged.depth = snakemake@input[[2]]

} else {

  # Data from lofreq and varscan labeled with subclonal haplotypes
  updated.data = "../../results/variants/validated_variants.csv"

  # Depth per position
  merged.depth = "../../results/coverage/merged.depth"

}

```

```{r Outputs, echo=T}

## ==== File paths output data ==== ##

if (exists("snakemake")) {
  
  # Data formatted for MACHINA/SPRUCE
  machina.input.data = snakemake@params[["spruce"]]

} else {

  # Data formatted for MACHINA/SPRUCE
  machina.input.data = "../../results/variants/spruce_input.tsv"

}

```

```{r Load Data, echo = T}

# Read in the variants with G1, G2, and G1-1 labeled
labeled.df = read_csv(updated.data, show_col_types = FALSE)

# Get the haplotype names
haplotypes.label = labeled.df %>% 
  select(SNP, Haplotype, Haplotype_Name, Background) %>% 
  distinct()

# Expand the mutations to have a frequency for every tissue.
expanded.df = labeled.df %>% 
  select(SNP, Tissue, AF) %>% 
  pivot_wider(names_from = "Tissue", values_from = "AF", values_fill = 0) %>% 
  pivot_longer(cols = !SNP, names_to = "Tissue", values_to = "AF") %>% 
  left_join(., select(labeled.df, c("SNP", "Tissue", "DP")), by = c("SNP", "Tissue")) %>% 
  mutate(DP = if_else(is.na(DP), 0, DP)) %>% 
  left_join(., haplotypes.label, by = "SNP") 

# Get the mean frequency for each haplotype
haplotype.mean = expanded.df %>% 
  group_by(Haplotype_Name, Tissue, Background) %>% 
  summarize(AF = mean(AF))

```

## Frequency by Compartment

One question is how the sum of frequencies of each haplotype looks across the tissue specimens. These shouldn't sum to more than the average frequency of their lineage (`Genome 1` or `Genome 2`). If they do, that means that some of these haplotypes could be descended from one another. 

```{r Color Palletes, echo=F}

# Make a color scheme for the haplotypes
## Genome-1
genome.1.haplotypes = c( "cluster 1", "cluster 2", "cluster 3", "cluster 4", "cluster 5", "cluster 7", "cluster 8", "genome 1", "genome 01")
genome.1.colors = brewer.pal(length(genome.1.haplotypes), "Blues")
names(genome.1.colors) = genome.1.haplotypes

## Genome-2
genome.2.haplotypes = c("cluster 9", "cluster 10", "cluster 11", "cluster 12", "cluster 13", "genome 2")
genome.2.colors = brewer.pal(length(genome.2.haplotypes), "YlOrRd")
names(genome.2.colors) = genome.2.haplotypes

## Cluster 6
cluster.6.haplotype = c("cluster 6")
cluster.6.color = brewer.pal(8, "Purples")[8]
names(cluster.6.color) = cluster.6.haplotype

## Combine
haplotype.colors = c(genome.1.colors, genome.2.colors, cluster.6.color)
# names(haplotype.colors) = gsub(" ", "_", names(haplotype.colors ))
haplotype.colors = c(haplotype.colors , "Anc" = "black")
# cluster 1 is too light
haplotype.colors['cluster 1'] = "#bdbebf"

## Set the Compartment Colors 
# Frontal Cortex
FCs = c("Frontal Cortex 1", "Frontal Cortex 2", "Frontal Cortex 3")
FCs.colors = brewer.pal(length(FCs), "Greens")
names(FCs.colors) = FCs
# Internal Compartments
Internal = c("Cerebellum Nucleus", "Cerebellum", "Midbrain", "Internal Capsule", "Brain Stem", "UBS", "Hippocampus")
Internal.colors = brewer.pal(length(Internal), "YlOrRd")
names(Internal.colors) = Internal
# Other Lobes
Other =  c("Occipital Lobe", "Temporal Lobe", "Parietal Lobe")
Other.colors = brewer.pal(length(Other), "Blues")
names(Other.colors) = Other
# Combine
Compartment.cols = c(FCs.colors, Internal.colors, Other.colors)
names(Compartment.cols) = gsub(" ", "_", names(Compartment.cols))

```

```{r Haplotype and Tissue Order, echo=F}
# Tissue order - roughly by the location in the brain 
expanded_tissue_order = c(
  "SSPE 1", 
  "SSPE 2",
  "Frontal Cortex 2", 
  "Frontal Cortex 1", 
  "Frontal Cortex 3", 
  "Parietal Lobe",
  "Temporal Lobe", 
  "Occipital Lobe",
  "Hippocampus",
  "Internal Capsule", 
  "Midbrain", 
  "Upper Brain Stem",
  "Brain Stem",
  "Cerebellum",
  "Cerebellum Nucleus"
)

haplotype.order = c("cluster 1",
                    "cluster 2",
                    "cluster 3",
                    "cluster 4",
                    "cluster 5",
                    "cluster 6",
                    "cluster 7",
                    "cluster 8",
                    "cluster 9",
                    "cluster 10",
                    "cluster 11",
                    "cluster 12",
                    "cluster 13",
                    "cluster 14")
```


### Genome 2

What does this look like for Genome 2?

```{r Genome 2 Clusters, warning=F, fig.align='center', fig.width=8, fig.height=5, echo=T}

genome.2.clusters = haplotype.mean %>% 
  filter(Background == "genome-2" & (!Haplotype_Name %in% c("genome 2"))) %>% 
  pull(Haplotype_Name) %>% 
  unique()

genome.2.mean = haplotype.mean %>% 
  filter(Haplotype_Name == "genome 2") %>% 
    mutate(Tissue = case_when(
      Tissue == "UBS" ~ "Upper Brain Stem",
      TRUE ~ Tissue
    ))

haplotype.mean %>%     
  mutate(Tissue = case_when(
      Tissue == "UBS" ~ "Upper Brain Stem",
      TRUE ~ Tissue
    )) %>% 
    filter(Haplotype_Name %in% genome.2.clusters) %>% 
    ggplot(aes(x = factor(Tissue, levels = expanded_tissue_order), y = AF, fill = (factor(Haplotype_Name, levels = haplotype.order)), group = Haplotype_Name)) +
      geom_area() +
      geom_line(data = genome.2.mean, aes(x = factor(Tissue, levels = expanded_tissue_order), col = (Haplotype_Name), y = AF, group = Haplotype_Name), linewidth = 2) +
      xlab("Tissue") + 
      ylab("Allele Frequency") +
      labs(fill="Genotype", col = "Cluster") +
      scale_fill_manual(values = haplotype.colors, name="") +
      scale_color_manual(values = haplotype.colors, guide = "none") +
      theme_bw(16) +
      theme(axis.text.x = element_text(angle = 35, hjust=1)) + 
      theme(strip.text.x = element_text(size = 12))

```

Genome 2 doesn't have any obvious violations! In fact, the sum of haplotypes that we think are descended from Genome 2 adds up to nearly the average frequency of Genome 2 in all 15 specimens. This suggests that there isn't much evidence based on frequency that these haplotypes are related to one another beyond being directly descended from Genome 2. 

### Genome 1

What about Genome 1?

```{r Genome 1 Clusters, warning=F, fig.align='center', fig.width=8, fig.height=5, echo=T}

genome.1.clusters = haplotype.mean %>% 
  filter(Background == "genome-1" & (!Haplotype_Name %in% c("genome 1", "genome 01"))) %>% 
  pull(Haplotype_Name) %>% 
  unique()

genome.1.mean = haplotype.mean %>% 
  filter(Haplotype_Name == "genome 01") %>% 
    mutate(Tissue = case_when(
      Tissue == "UBS" ~ "Upper Brain Stem",
      TRUE ~ Tissue
    ))

haplotype.mean %>% 
  select(!Background) %>% 
  filter(Haplotype_Name != "subclonal") %>% 
    mutate(Tissue = case_when(
      Tissue == "UBS" ~ "Upper Brain Stem",
      TRUE ~ Tissue
    )) %>% 
    # Cluster 3 is on the background cluster 4
  pivot_wider(names_from = Haplotype_Name, values_from = AF) %>%
  pivot_longer(cols = -c(Tissue), names_to = "Haplotype_Name", values_to = "AF") %>% 
  filter(Haplotype_Name %in% genome.1.clusters) %>% 
    ggplot(aes(x = factor(Tissue, levels = expanded_tissue_order), y = AF, fill = (factor(Haplotype_Name, levels = haplotype.order)), group = Haplotype_Name)) +
      geom_area() +
      geom_line(data = genome.1.mean, aes(x = factor(Tissue, levels = expanded_tissue_order), col = (Haplotype_Name), y = AF, group = Haplotype_Name), linewidth = 2) +
      xlab("Tissue") + 
      ylab("Allele Frequency") +
      labs(fill="Genotype", col = "Cluster") +
      scale_fill_manual(values = haplotype.colors, name="") +
      scale_color_manual(values = haplotype.colors, guide = "none") +
      theme_bw(16) +
      theme(axis.text.x = element_text(angle = 35, hjust=1)) + 
      theme(strip.text.x = element_text(size = 12))

```

In Genome 1 there is a violation in SSPE 2 where two clusters sum to a frequency greater than the frequency of Genome 1 in that specimen. This gives us reason to suspect that these clusters are related in some way. In particular, that cluster 3 is on the background of cluster 4. So, I'll subtract the average allele frequency of cluster 3 from cluster 4 in every compartment and see what this looks like. 

```{r All Clusters, warning=F, fig.align='center', fig.width=8, fig.height=10, echo=T}

haplotype.clusters = haplotype.mean %>% 
  filter(Background %in% c("genome-1", "genome-2") & (!Haplotype_Name %in% c("genome 1", "genome 01", "genome 2"))) %>% 
  pull(Haplotype_Name) %>% 
  unique()

genome.means = haplotype.mean %>% 
  filter(Haplotype_Name %in% c("genome 01", "genome 2")) %>% 
    mutate(Tissue = case_when(
      Tissue == "UBS" ~ "Upper Brain Stem",
      TRUE ~ Tissue
    )) %>% 
  mutate(Background = case_when(
    Background == "genome-1" ~ "Genome 1",
    Background == "genome-2" ~ "Genome 2",
    TRUE ~ Background
  ))

backgrounds = haplotype.mean %>% 
  ungroup() %>% 
  select(Haplotype_Name, Background) %>% 
  distinct() %>% 
  mutate(Background = case_when(
    Background == "genome-1" ~ "Genome 1",
    Background == "genome-2" ~ "Genome 2",
    TRUE ~ Background
  ))

haplotype.mean %>% 
  select(!Background) %>% 
  filter(!Haplotype_Name %in% c("subclonal", "both")) %>% 
    mutate(Tissue = case_when(
      Tissue == "UBS" ~ "Upper Brain Stem",
      TRUE ~ Tissue
    )) %>% 
  pivot_wider(names_from = Haplotype_Name, values_from = AF) %>%
  # Cluster 3 is on the background cluster 4 on Genome 01
  mutate(`cluster 4` = `cluster 4` - `cluster 3`) %>% 
  # If the average frequency is below 0, return it to 0
  mutate(`cluster 4` = if_else(`cluster 4` < 0, 0, `cluster 4`)) %>% 
  pivot_longer(cols = -c(Tissue), names_to = "Haplotype_Name", values_to = "AF") %>% 
  filter(Haplotype_Name %in% haplotype.clusters) %>% 
  left_join(., backgrounds, by = "Haplotype_Name") %>% 
    ggplot(aes(x = factor(Tissue, levels = expanded_tissue_order), y = AF, fill = (factor(Haplotype_Name, levels = haplotype.order)), group = Haplotype_Name)) +
      geom_area() +
      geom_line(data = genome.means, aes(x = factor(Tissue, levels = expanded_tissue_order), col = (Haplotype_Name), y = AF, group = Haplotype_Name), linewidth = 2) +
      facet_wrap(~Background, ncol = 1) +
      xlab("Tissue") + 
      ylab("Allele Frequency") +
      labs(fill="Genotype", col = "Cluster") +
      scale_fill_manual(values = haplotype.colors, name="", breaks=haplotype.order) +
      scale_color_manual(values = haplotype.colors, guide = "none") +
      theme_bw(16) +
      theme(axis.text.x = element_text(angle = 35, hjust=1)) + 
      theme(strip.text.x = element_text(size = 12))

ggsave("./figure-5bc.png", width = 8, height = 10)

```
However, we want to confirm these relationships as best as possible using a systematic approach with the algorithm SPRUCE. Using SPRUCE, we can fit trees to the data based on haplotype frequency. Then, we can use bridging reads to filter out trees that link haplotypes that we know must be forbidden based on the overlapping reads. 

## Prepare data for Spruce

But first, we need to format the input data for SPRUCE. The naive approach is to use the min and max allele frequency of SNPs in each compartment for each haplotype as the lower and upper bounds. 

```{r Format Data for SPRUCE, message=F, warning=F, fig.align='center', fig.width=19, fig.height=10, echo=T}

# Add in the depth data
merged.depth.df = fread(merged.depth) %>% 
  select(!V1)

# Join the depth data to the allele frequencies
expanded.depth.df = expanded.df %>% 
  select(!DP) %>% 
  left_join(., distinct(select(labeled.df, POS, SNP, Tissue, Accession)), by = c("SNP", "Tissue")) %>% 
  left_join(., merged.depth.df, by = c("POS", "Accession")) %>% 
  # Rename the UBS
  mutate(Tissue = if_else(Tissue == "UBS", "Upper Brain Stem", Tissue))

# Amend C4573T in the Cerebellum to be the mean haplotype frequency
g1.cerrebellum.mean = filter(haplotype.mean, Haplotype_Name == 'genome 01', Tissue == 'Cerebellum')$AF
expanded.depth.df = expanded.depth.df %>% 
  mutate(AF = case_when(
    SNP == "C4573T" & Tissue == 'Cerebellum' ~ g1.cerrebellum.mean,
    TRUE ~ AF
  )) 

# Make tissue index
Tissue_index = expanded.depth.df %>% 
  select(Tissue) %>% 
  distinct() %>% 
  mutate(tind = 1:n() - 1)

# Make cluster index 
Cluster_index = expanded.depth.df %>% 
  select(Haplotype_Name) %>% 
  distinct() %>% 
  filter(!Haplotype_Name %in% c("both", "subclonal")) %>% 
  mutate(cind = 1:n() - 1) %>%
  add_row(Haplotype_Name = "Anc", cind = 16) 

# Calculate the MAF along with some variance (in this case the min and max in each compartment) for wiggle room. 
spruce.data.df = expanded.depth.df %>%
  filter(!Haplotype_Name %in% c("both", "subclonal")) %>% 
  group_by(Haplotype_Name, Tissue) %>% 
  # use this formulation for cluster allele frequency, because it puts less weight on low read depth positions
  summarize(MAF = mean(AF),
            min = min(AF), 
            max = max(AF)) %>% 
  # MACHINA wants an upper and lower bound on what the allele frequencies are
  mutate(lb = pmax(0, min), ub = pmin(1, max))
 
spruce.data.df %>% 
  ggplot(aes(x = factor(Tissue, levels = expanded_tissue_order), y = MAF, group = Haplotype_Name, ymin=lb, ymax=ub)) + 
    geom_ribbon(alpha=0.2, colour = "grey") +
    geom_line() +
    facet_wrap(~Haplotype_Name) + 
    xlab("Tissue") + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

Generally, this approach isn't so optimal. A better way might be to apply some downscale factor to depth and then use depth to calculate the variance. Let's try this below.

```{r Alternative Format Data for SPRUCE, message=F, warning=F, fig.align='center', fig.width=19, fig.height=10, echo=T}

down.wieght = 0.0005

spruce.data.alt.df = expanded.df %>%
  group_by(Haplotype_Name, Tissue) %>%
  filter(!Haplotype_Name %in% c("both", "subclonal")) %>%
  # Add a psuedo-count to depth 
  mutate(DP = DP + 1) %>% 
  # Add back the uncovered G1 SNP
  mutate(AF = case_when(
    SNP == "C4573T" & Tissue == 'Cerebellum' ~ g1.cerrebellum.mean,
    TRUE ~ AF
  )) %>% 
  # Rename the UBS
  mutate(Tissue = if_else(Tissue == "UBS", "Upper Brain Stem", Tissue)) %>% 
  # use this formulation for cluster allele frequency, because it puts less weight on low read depth positions
  summarize(MAF = sum(AF * DP)/sum(DP),
            var = sqrt(MAF * (1 - MAF) / (sum(DP) * down.wieght))) %>%
  # MACHINA wants an upper and lower bound on what the allele frequencies are
  mutate(lb = pmax(0, MAF - var), ub = pmin(1, MAF + var))


spruce.data.alt.df %>% 
  ggplot(aes(x = factor(Tissue, levels = expanded_tissue_order), y = MAF, group = Haplotype_Name, ymin=lb, ymax=ub)) + 
    geom_ribbon(alpha=0.2, colour = "grey") +
    geom_line() +
    facet_wrap(~Haplotype_Name) + 
    xlab("Tissue") + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  

```

Currently, it's not clear exactly which method is optimal. One approach takes into account the depth in the tissue to get variance, the other takes into account the variance in freuency in each tissue. We'll stick with the latter fequency-based approach for now. 

```{r Write out Data}

# Add the ancestor back in or it won't connect g1 and g2
with.anc.df = bind_rows(spruce.data.df, 
                    Tissue_index %>%
                      select(-tind) %>% 
                      mutate(Haplotype_Name = "Anc", MAF = NA, SD = NA, lb = 1, ub = 1)
                    )

# Merge the average cluster frequencies with the index information that spruce wants
spruce.output = left_join(left_join(with.anc.df, Tissue_index), Cluster_index) %>% 
    ungroup() %>%
    # need to substitute out spaces
    mutate(Tissue = gsub(" ", "_", Tissue)) %>%
    mutate(Haplotype = gsub(" ", "_", Haplotype_Name)) %>%
    mutate(sind = tind, sample = Tissue) %>%
    select(tind, Tissue, sind, sample, cind, Haplotype, lb, ub)  %>% 
    arrange(tind, cind) %>% 
    mutate(ub = pmax(lb, ub)) 

head(spruce.output) 

```

```{r Write out Machina inputs}

## ==== Write Output data ==== ##

# Header needs to specify these variables
write(
  paste0(
    nrow(Tissue_index),
    "\n", nrow(Tissue_index),
    "\n", nrow(Cluster_index),
    "\n#", collapse = "" ),
  machina.input.data
)

# Then write the cluster frequencies 
write.table(spruce.output,
            machina.input.data,
            sep = "\t", 
            quote = FALSE, 
            row.names = FALSE,
            col.names = FALSE,
            append = TRUE)

# Run this command
paste0("generatemutationtrees ", machina.input.data, " > ", "../../results/variants/spruce_output.tsv")
```

